Thesis Title

### 2022-03-03 chapters/introduction.tex
BAHH

Because this approach is PCR-based, it suffers from the same issues described previously for conventional PCR, requiring primer design. Additionally, it must necessarily account for intragenomic variation between 16S gene copies. Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 bp) gene \citep{johnson_evaluation_2019}. The emergence of third generating sequencing technologies (see \secref{ssec:3rd_gen_seq}) allows for this limitation to be overcome but currently, only with only a fraction of the databases include complete 16S rRNA sequences.

While viruses are an integral part of the microbiota, no universal viral marker genes are available to perform such taxonomic assignments. Amplification of whole viral genomes is possible and, in 2015, RNA extracted from whole blood, serum, re-suspended swabs and urine, after targeted amplification of the whole viral genome, proved invaluable in the track of the Ebola virus disease epidemic in West Africa, responsible for >11 thousands deaths, allowing for the characterisation of the infectious agent the determination of its evolutionary rate, signatures of host adaptation, identification and monitoring of diagnostic targets and responses to vaccines and treatments \citep{quick_real-time_2016}. As an alternative, broad scope viral targeted sequence capture (TSC) panels offer depletion of background nucleic acids and improve the recovery of viral reads by targeting coding sequence from a multitude viral genera, such as VirCapSeq-VERT Capture Panel\footnote{\url{https://sequencing.roche.com/content/dam/rochesequence/worldwide/resources/brochure-vircapseq-vert-capture-panel-SEQ1000117.pdf}} but do not guarantee the full recovery of the viral genome, and can present biases towards certain genera \citep{schuele_assessment_2020, wylie_enhanced_2015}.

@@

Clinical applications of shotgun metagenomics derives its roots from the use of microarrays (see \secref{ssec:current_standards}), where is was successfully applied in in-depth microbiome analysis of different sites in the human body, it was the emergence of second-generation sequencing technology and its high throughput of genomic data at a competitive price that made the sequencing of all genomic content, DNA and/or RNA) if a clinical sample a viable possibility for diagnostics (see \secref{ssec:2nd_gen_seq}) \citep{miller_basic_2009, palmer_rapid_2006, chiu_clinical_2019}. The first reported case that demonstrated the utility of shotgun metagenomics was in 2014 with the clinical diagnosis of neuroleptospirosis in a 14-year-old immunodeficient and critically ill boy with meningoencephalitis by \cite{wilson_actionable_2014}, prompting appropriate targeted antibiotic treatment and eventual recovery of the patient. In this case, traditional methods, including an invasive brain biopsy, failed to provide answers, until the shotgun sequencing of cerebrospinal fluid identified 475 of 3,063,784 sequence reads (0.016\%) corresponding to leptospira, for which clinical assays were negative due to its very low abundance. Ever since, many other reports of successful application of shotgun metagenomics in clinical metagenomics have been reported. but all in edge cases where traditional diagnostic methods have failed or as proof-of-concept \citep{couto_critical_2018, vijayvargiya_application_2019, sanabria_shotgun-metagenomics_2020, hirakata_application_2021}.

@@

Clinical diagnostic applications have lagged behind research advances. A significant challenge with shotgun metagenomic approach is the large variation in the pathogen load between patient samples, as evidenced in the studies presented. A low pathogen load and a high contamination of host DNA or even the present microbiome may result in enough data to produce the high-resolution subtype needed to distinguish and cluster the cases that were caused by the same outbreak pathogen source, or, extremely, the undetection of the causative agent \citep{carleton_metagenomic_2019, chiu_clinical_2019}. Differential lysis of human host cells followed by degradation of background DNA has proven an effective method to reduce host contamination, but limitations include potential decreased sensitivity for microorganisms without cell walls, such as \textit{Mycoplasma} spp. or parasites; a possible paradoxical increase in exogenous background contamination by use of additional reagent \citep{salter_reagent_2014, oneil_ribosomal_2013, feehery_method_2013}. Additionally, it is often unclear whether a detected microorganism is a contaminant, coloniser or \textit{bona fide} pathogen, and with the lack of golden standards remains one of the biggest challenges when applying this methods in clinical microbiology for diagnosis.

In addition to negative controls, already a common practise in any sequencing assay and in particular in metataxonomics (see \secref{sssec:metataxonomics}), positive controls can be a way to circumvent the lack of golden standards, either through the spike of the samples with a known amount of a specific DNA/RNA or though the sequencing of samples with a known composition and abundance. Well-characterised reference standards and controls are needed to ensure shotgun metagenomics assay quality and stability over time \citep{chiu_clinical_2019, mcintyre_comprehensive_2017}. Most available metagenomic reference materials are highly tailored to a specific application. For example, the ZymoBIOMICS Microbial Community Standard\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}} is the first commercially available standard for microbiomics and metagenomics studies, providing mock a mock community with defined composition and abundance consisting of Gram-positive, Gram-negative and yeast. It is useful to determine the limit of detection of an assay, and the effectiveness and biases of a given protocol. Standards with a more limited spectrum of organisms are also available, such as the National Institute of Standards and Technology (NIST)\footnote{\url{https://www.nist.gov/}} reference materials for mixed microbial DNA detection, which contain only bacteria. Thus, these materials may not be applicable to untargeted shotgun metagenomics analyses.

@@

As stated previously (see \secref{sec:genomics_approach} and \secref{ssec:metagenomics}), one of the biggest challenges when dealing with genomic, and in particular metagenomic, data is the lack of golden standards. This is also applicable to the bioinformatic analysis, required due to the amount of data produced by genomic sequencing technologies. This is currently one of the bottleneck in the deployment of sequencing technology in clinical microbiology as there's no standard in how to deal with the increasing amount of data produced in a fit-for-purpose manner \citep{carrico_primer_2018}. Bioinformatics is an interdisciplinary research field that applies methodologies from computer science, applied mathematics and statistics to the study of biological phenomena\citep{carrico_primer_2018}. With the widespread use and continuous development of sequencing technologies, bioinformatics has become a cornerstone in modern clinical microbiology.

@@

All sequencing technologies, regardless of generation, produce data in the the same standard file format: the FASTQ, a text-based format for storing both a biological sequence (usually nucleotide sequence) and its corresponding quality scores \citep{cock_sanger_2010}. Originally developed at the Wellcome Trust Sanger Institute, the FASTQ has emerged as a common file format for sharing sequencing read data (see \ref{fig:figure3}). The FASTQ can be considered as an extension of the ‘FASTA sequence file format’, originally invented by \cite{pearson_improved_1988}, which includes just the sequence information. A FASTQ file normally uses four lines per sequence:

@@

@@

An important information that can be retrieved directly from the quality controlled read data is the relative abundance of the detected organisms. Kraken (Wood & Salzberg, 2014) and Bracken (Lu et al., 2017), Midas (Nayfach et al., 2016), Kaiju (Menzel et al., 2016) and MetaPhlAn 2 (Truong et al., 2015) are all examples of taxonomic classifiers for short-read raw sequencing data that provide information on relative abundance. All tools provide a reference database. Kraken offers several pre-built databases constructed from complete bacterial, archaeal, and viral genomes in RefSeq (as of Oct. 18, 2017), known as MiniKraken, that range in size to accommodate limited computational resources, as well as providing the option to build the standard database given that the required computational resources are met. Midas and MetaPhlAn also provide a default pre-built database, with resolution down to strain level. Kaiju differs from the other tools by using a protein reference database but no pre-built version is available, requiring significant resources to build and index the database pre-use. 
The long-read data can be treated as single-end fastq, after proper conversion in the basecalling process. All tools mentioned accommodate classification of single-end files. 
16S classification can also be performed, for validation purposes, by extracting the reads of interest through HMMER (Wheeler & Eddy, 2013) and then perform a traditional OTU analysis with Quiime2 (https://qiime2.org/) or MapSEQ (Rodrigues et al., 2017).

@@

% TRASH TEXT
A plethora of open-source tools are available specifically for metagenomic data, both short and long-read data, and several combinations of these tools can be used to characterize the causative agent in a patient's infection in a fraction of the time required by the traditional methods. Alongside, there are several commercial alternatives, such as CLC Genomics Workbench (QIAGEN Bioinformatics), Taxonomer (Flygare et al., 2016) and BaseSpace (Illumina), that offer ready to use complete workflows at the cost of lack of transparency, reproducibility and control in the analysis.
Several steps that can be implemented to ensure the transparency and reproducibility of the chosen workflow. Favouring open-source tools, with clear documentation describing the methodology implemented, and stating the version of the software used and which parameters were used enables the comparison of results. This can be simplified by containerizing all the software tools with one of the many solutions available, like Docker (https://www.docker.com/) or Singularity (Kurtzer et al., 2017). The use of workflow managers, like nextflow (Tommaso et al., 2017) or the Galaxy Project (Afgan et al., 2016), will push reproducibility to the next level by taking advantage of the containerization and scalability, enabling the workflow to be executed with the same parameters in the same conditions in a multitude of different environments. The FlowCraft project (https://github.com/assemblerflow/flowcraft) leverages the combination of Nextflow and docker/singularity containers to assemble, monitor and report scientific pipelines created from the combination of pre-built components, many of them supporting metagenomic analysis.
Additional difficulties of metagenomic data are the overpowering quantities of host DNA that are often sequenced (Couto et al., 2018), making the microbial community close to undetectable, the presence of contaminants, from the bench process to the biota, and the cost associated with this methodology. They account for major caveats and must be made aware of when analysing the data.
The basic strategies for analysing metagenomic data can be simplified in the scheme in Figure 1. One of the biggest challenges when doing metagenomic analysis is differentiating between colonization and infection and to successfully discriminate between a potential pathogen and background microbiota. In the latter, when analysing samples from presumably sterile sites, like CSF and blood, it is safe to assume that all organisms found are of interest. In locations with a microbiota, the use of spiked metagenomic samples as positive control might guide the detection of the possible pathogens by comparing relative abundance between the samples. The inclusion of negative controls is essential for the correct identification of contaminants in the taxonomic results, whether originated from the sample collection, handling or sequencing process. These controls should be processed similarly to the samples and the taxonomic results should be filtered out from the final reports.

>>>>

Because this approach is PCR-based, it suffers from the same issues described previously for conventional PCR, requiring primer design. Additionally, it must necessarily account for intragenomic variation between 16S gene copies. Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 bp) gene \citep{johnson_evaluation_2019}. The emergence of third generating sequencing technologies (see \secref{ssec:3rd_gen_seq}) allows for this limitation to be overcome but currently,  only a fraction of the databases includes complete 16S rRNA sequences.

While viruses are an integral part of the microbiota, no universal viral marker genes are available to perform such taxonomic assignments. Amplification of whole viral genomes is possible and, in 2015, RNA extracted from whole blood, serum, re-suspended swabs and urine, after targeted amplification of the whole viral genome, proved invaluable in the track of the Ebola virus disease epidemic in West Africa, responsible for >11 thousand deaths, allowing for the characterisation of the infectious agent the determination of its evolutionary rate, signatures of host adaptation, identification and monitoring of diagnostic targets and responses to vaccines and treatments \citep{quick_real-time_2016}. As an alternative, broad scope viral targeted sequence capture (TSC) panels offer depletion of background nucleic acids and improve the recovery of viral reads by targeting coding sequence from a multitude viral genera, such as VirCapSeq-VERT Capture Panel\footnote{\url{https://sequencing.roche.com/content/dam/rochesequence/worldwide/resources/brochure-vircapseq-vert-capture-panel-SEQ1000117.pdf}} but do not guarantee the full recovery of the viral genome, and can present biases towards certain genera \citep{schuele_assessment_2020, wylie_enhanced_2015}.

@@

Clinical applications of shotgun metagenomics derive its roots from the use of microarrays (see \secref{ssec:current_standards}), where it was successfully applied in in-depth microbiome analysis of different sites in the human body, it was the emergence of second-generation sequencing technology and its high throughput of genomic data at a competitive price that made the sequencing of all genomic content, DNA and/or RNA) if a clinical sample a viable possibility for diagnostics (see \secref{ssec:2nd_gen_seq}) \citep{miller_basic_2009, palmer_rapid_2006, chiu_clinical_2019}. The first reported case that demonstrated the utility of shotgun metagenomics was in 2014 with the clinical diagnosis of neuroleptospirosis in a 14-year-old immunodeficient and critically ill boy with meningoencephalitis by \cite{wilson_actionable_2014}, prompting appropriate targeted antibiotic treatment and eventual recovery of the patient. In this case, traditional methods, including an invasive brain biopsy, failed to provide answers, until the shotgun sequencing of cerebrospinal fluid identified 475 of 3,063,784 sequence reads (0.016\%) corresponding to leptospira, for which clinical assays were negative due to its very low abundance. Ever since many other reports of successful application of shotgun metagenomics in clinical metagenomics have been reported. but all in edge cases where traditional diagnostic methods have failed or as proof-of-concept \citep{couto_critical_2018, vijayvargiya_application_2019, sanabria_shotgun-metagenomics_2020, hirakata_application_2021}.

@@

Clinical diagnostic applications have lagged behind research advances. A significant challenge with shotgun metagenomic approach is the large variation in the pathogen load between patient samples, as evidenced in the studies presented. A low pathogen load and  high contamination of host DNA or even the present microbiome may result in enough data to produce the high-resolution subtype needed to distinguish and cluster the cases that were caused by the same outbreak pathogen source, or, extremely, the undetection of the causative agent \citep{carleton_metagenomic_2019, chiu_clinical_2019}. Differential lysis of human host cells followed by degradation of background DNA has proven an effective method to reduce host contamination, but limitations include potential decreased sensitivity for microorganisms without cell walls, such as \textit{Mycoplasma} spp. or parasites; a possible paradoxical increase in exogenous background contamination by use of additional reagent \citep{salter_reagent_2014, oneil_ribosomal_2013, feehery_method_2013}. Additionally, it is often unclear whether a detected microorganism is a contaminant, coloniser or \textit{bona fide} pathogen, and the lack of golden standards remains one of the biggest challenges when applying these methods in clinical microbiology for diagnosis.

In addition to negative controls, already a common practice in any sequencing assay and in particular in metataxonomics (see \secref{sssec:metataxonomics}), positive controls can be a way to circumvent the lack of golden standards, either through the spike of the samples with a known amount of a specific DNA/RNA or though the sequencing of samples with known composition and abundance. Well-characterised reference standards and controls are needed to ensure shotgun metagenomics assay quality and stability over time \citep{chiu_clinical_2019, mcintyre_comprehensive_2017}. Most available metagenomic reference materials are highly tailored to a specific application. For example, the ZymoBIOMICS Microbial Community Standard\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}} is the first commercially available standard for microbiomics and metagenomics studies, providing mock a mock community with defined composition and abundance consisting of Gram-positive, Gram-negative and yeast. It is useful to determine the limit of detection of an assay, and the effectiveness and biases of a given protocol. Standards with a more limited spectrum of organisms are also available, such as the National Institute of Standards and Technology (NIST)\footnote{\url{https://www.nist.gov/}} reference materials for mixed microbial DNA detection, which contain only bacteria. Thus, these materials may not apply to untargeted shotgun metagenomics analyses.

@@

As stated previously (see \secref{sec:genomics_approach} and \secref{ssec:metagenomics}), one of the biggest challenges when dealing with genomic, and in particular metagenomic, data is the lack of golden standards. This is also applicable to the bioinformatic analysis, required due to the amount of data produced by genomic sequencing technologies. This is currently one of the bottlenecks in the deployment of sequencing technology in clinical microbiology as there's no standard in how to deal with the increasing amount of data produced in a fit-for-purpose manner \citep{carrico_primer_2018}.

Bioinformatics is an interdisciplinary research field that applies methodologies from computer science, applied mathematics and statistics to the study of biological phenomena\citep{carrico_primer_2018}. With the widespread use and continuous development of sequencing technologies, bioinformatics has become a cornerstone in modern clinical microbiology.

@@

In all sequencing technologies (see \secref{ssec:sequencing}), many copies of the source DNA are randomly fragmented and sequenced. To these sequences, we refer to as reads. In the case of second-generation sequencing (see \secref{ssec:2nd_gen_seq}), one or both ends of the fragment can be sequenced. If a fragment is sequenced from one end, we refer to it as single-end sequencing. If a fragment is sequenced on both ends, spanning the entire fragment, it is called paired-end sequencing.

All sequencing technologies, regardless of generation, produce data in the same standard file format: the FASTQ, a text-based format for storing both a biological sequence (usually nucleotide sequence) and its corresponding quality scores \citep{cock_sanger_2010}. Originally developed at the Wellcome Trust Sanger Institute, the FASTQ has emerged as a common file format for sharing sequencing read data (see \ref{fig:figure3}). The FASTQ can be considered as an extension of the ‘FASTA sequence file format’, originally invented by \cite{pearson_improved_1988}, which includes just the sequence information. A FASTQ file normally uses four lines per sequence:

@@

With the lack of golden standards for metagenomic analysis, the use of simulated mock communities, with known composition, abundance and genomic information, provides a ground truth against which evaluations of success can be made. Given their standard structure and adoption, the generation of simulated FASTQ files from a reference, or a set of references, is very straightforward.

Multiple computational tools for the simulation of sequencing data, particularly for second and third-generation sequencing technologies, have been developed in recent years, which could be used to compare existing and new bioinformatic analytical pipelines. \cite{escalona_comparison_2016} provides a comprehensive assessment of 23 different read-simulation tools,  highlighting their distinct functionality, requirements and potential applications, as well as providing a selection of suggestions for different simulation tools depending on their purpose. For \textit{in silico} metagenomic sequence generation, a total of XX tools are available for second and third-generation reads (see \ref{}).

% ADD FIGURE!

@@

A piece of important information that can be retrieved directly from the quality-controlled read data: (1) the identification and characterisation of the microbes present in a sample and (2) their relative abundance. Taxonomic classification methods can vary depending on the sequencing methodology used: pure culture, metataxonomics and amplicon metagenomics, and shotgun metagenomics.

From pure culture, taxonomic identification of the read content of a sample is useful to assess contamination. Tools like Kraken2 \citep{wood_kraken_2014, wood_improved_2019} and Braken \citep{lu_bracken_2017}. These tools, relying on a database, assign taxonomic labels to reads and are therefore biased to the contents of the database used. Various databases are available\footnote{\url{https://benlangmead.github.io/aws-indexes/k2}}, varying in size and content (archaea, bacteria, viral, plasmid, human and eukaryotic pathogens), and therefore in sensitivity depending on the resources available and the purpose intended. Alternatively, there are options to create custom databases.

These tools are also extremely useful to assess the contents of a metagenomic sample. Alternatives such as Midas \citep{nayfach_integrated_2016}, Kaiju, \citep{menzel_fast_2016}, and MetaPhlAn2 \citep{} offer the same analysis as Kraken and Bracken using different algorithms, and with the disadvantage that they come prepackaged with their own databases, without the option to create a tailored database, limiting their applicability. Kaiju differs from the other tools by using a protein reference database, instead of nucleotide, but no pre-built version is available, requiring significant resources to build and index the database pre-use. The long-read data of third-generation sequencing technologies (see \secref{ssec:3rd_gen_seq}) can be treated as single-end reads, and all tools mentioned accommodate the classification of single-end files.

\subsubsection{Metataxonomics and Operational Taxonomic Units}

Metataxonomics (see \secref{sssec:metataxonomics}) is the most widely used technique for microbial diversity analysis \citep{hilton_metataxonomic_2016}, and due to its particularities, the analysis of this data is also very particular.

The same approach of direct taxonomic classification can be followed as with genomic and shotgun metagenomic data, given that the databases include rRNA sequences, but more traditional approaches rely on the identification of OTUs. This is done through sequence matching with the appropriate databases  (see \secref{sssec:metataxonomics}), followed by statistical analysis to assess differences between given groups. Quiime2\footnote{\url{https://qiime2.org/}} \citep{bolyen_reproducible_2019} has become the \textit{de facto} tool for metataxonomic analysis as a framework with an ever-growing suite of plugins and intuitive data visualisation tools for the assessment of results.

@@

% TRASH TEXT

s
A plethora of open-source tools are available specifically for metagenomic data, both short and long-read data, and several combinations of these tools can be used to characterize the causative agent in a patient's infection in a fraction of the time required by the traditional methods. Alongside, there are several commercial alternatives, such as CLC Genomics Workbench (QIAGEN Bioinformatics), Taxonomer (Flygare et al., 2016) and BaseSpace (Illumina), that offer ready to use complete workflows at the cost of lack of transparency, reproducibility and control in the analysis.
Several steps that can be implemented to ensure the transparency and reproducibility of the chosen workflow. Favouring open-source tools, with clear documentation describing the methodology implemented, and stating the version of the software used and which parameters were used enables the comparison of results. This can be simplified by containerizing all the software tools with one of the many solutions available, like Docker (https://www.docker.com/) or Singularity (Kurtzer et al., 2017). The use of workflow managers, like nextflow (Tommaso et al., 2017) or the Galaxy Project (Afgan et al., 2016), will push reproducibility to the next level by taking advantage of the containerization and scalability, enabling the workflow to be executed with the same parameters in the same conditions in a multitude of different environments. The FlowCraft project (https://github.com/assemblerflow/flowcraft) leverages the combination of Nextflow and docker/singularity containers to assemble, monitor and report scientific pipelines created from the combination of pre-built components, many of them supporting metagenomic analysis.
Additional difficulties of metagenomic data are the overpowering quantities of host DNA that are often sequenced (Couto et al., 2018), making the microbial community close to undetectable, the presence of contaminants, from the bench process to the biota, and the cost associated with this methodology. They account for major caveats and must be made aware of when analysing the data.
The basic strategies for analysing metagenomic data can be simplified in the scheme in Figure 1. One of the biggest challenges when doing metagenomic analysis is differentiating between colonization and infection and to successfully discriminate between a potential pathogen and background microbiota. In the latter, when analysing samples from presumably sterile sites, like CSF and blood, it is safe to assume that all organisms found are of interest. In locations with a microbiota, the use of spiked metagenomic samples as positive control might guide the detection of the possible pathogens by comparing relative abundance between the samples. The inclusion of negative controls is essential for the correct identification of contaminants in the taxonomic results, whether originated from the sample collection, handling or sequencing process. These controls should be processed similarly to the samples and the taxonomic results should be filtered out from the final reports.
